---
title: "Estradiol analyses"
author: "Lev Kolinski"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    code_folding: hide
---

```{r setup, include=FALSE}
# loading packages
library(tidyverse)
library(lme4)
library(car)

# setting seed for reproducibility
set.seed(1234)

# Loading data
e_data<-read.csv("/Users/levkolinski/Desktop/taboga_estradiol/Data/estradiol_dataset.csv", sep = ";") 

e_data<-e_data |> 
  ## adding in age year column
  mutate(age_at_sample_years = as.numeric(sub("\\..*", "", age_at_sample))) |> 
  ## adding in hour column to account for diurnal variation
  mutate(hour = as.numeric(str_extract(time, "^[^:]+"))) 
  

duckface_data<-read.csv("/Users/levkolinski/Desktop/taboga_estradiol/Data/duckfaces_dataset.csv", sep = ";")

duckface_data$Date<-as.Date(duckface_data$Date)
duckface_data<-duckface_data |> 
  rename("ind_id" = Actor) |> 
  rename("date" = Date) |> 
  select(ind_id, date, age_at_sample)
```

# **Menarche and E**

## Age at menarche on the population-level

### Approach 1: Mean + 2SD

Adapting methods described in the poster uploaded on Google Drive, I first calculated the mean pre-pubertal fE2 concentration. Because infants/juvs have elevated fE2 when nursing, we need to exclude these individuals when determining what typical pre-pubertal fE2 looks like.

~~Per Slack communication on 7/17: defining prebuteral E from individuals greater than 2.5 years (def not nursing) and under 3 (def not pubertal)~~

Per Zoom meeting on 7/22: defining prepubertal E from individuals between 3 and 4 years old.

#### Step 1: Preparing dataset of prepubertal individuals (3 yo)

```{r,  class.source = 'fold-hide'}
## calculating average pre-pubertal E2 
prepubertal_data<-e_data |> 
  filter(age_at_sample >= 3.0 & age_at_sample <4) |>  # we want age 3..def pre menstruation
  select(ind_id, age_at_sample, e_conc_ug) 
```

#### Step 2: Calculating average prepubertal E

```{r}
prepubertal_e_data <- prepubertal_data |> 
   summarise(mean_e_conc_ug = mean(e_conc_ug, na.rm = TRUE),
            sd_e_conc_ug = sd(e_conc_ug, na.rm = TRUE)) 

mean_prepubertal_e_conc_ug <- prepubertal_e_data$mean_e_conc_ug
sd_prepubertal_e_conc_ug <- prepubertal_e_data$sd_e_conc_ug
```

#### Step 3: Calculating pubertal threshold E

Using the mean + 2SD approach:

```{r}
calculated_treshold <- mean_prepubertal_e_conc_ug + 2*sd_prepubertal_e_conc_ug

print(paste("Calculated fE2 threshold for menarche: ", calculated_treshold, " ug/g"))
```

#### Step 4: Calculating population average age at menarche based on threshold E

Looking at the average age at which the population has an E reading that crosses the threshold E. Restricting dataset to individuals who we started sampling pre menstruation (before age 3.75); only looking for

```{r}
# Identify the first age where e_conc_ug rises above the threshold for each individual
first_above_threshold <- e_data %>%
  group_by(ind_id) %>%
  filter(min(age_at_sample) <= 3.75) |> # want to make sure we have samples starting before menstruation to assign cutoff age 
  ungroup() |> 
  filter(age_at_sample>3.75) |> # only considering potentially pubescent if older than 4
  filter(e_conc_ug > calculated_treshold) %>%
  group_by(ind_id) %>%
  summarise(first_age_above_threshold = min(age_at_sample, na.rm = TRUE)) %>%
  ungroup()

# Calculate the mean and standard error of the ages
mean_age_above_threshold <- mean(first_above_threshold$first_age_above_threshold, na.rm = TRUE)
se_age_above_threshold <- sd(first_above_threshold$first_age_above_threshold, na.rm = TRUE) / sqrt(nrow(first_above_threshold))

print(paste("Average age at which e_conc_ug rises above the threshold: ", mean_age_above_threshold))
print(paste("Standard error of the average age: ", se_age_above_threshold))
```

### Plot of fE2 as a function of age: population-level

For interpretability, this plot just has the data sorted into age-years, rather than exact age estimations with lots of decimals

```{r, class.source = 'fold-hide'}
e_data_for_menarche <- e_data %>%
  filter(age_at_sample<8) |> 
  mutate(age_group = cut(age_at_sample, breaks = seq(0, max(age_at_sample, na.rm = TRUE) + 1, by = 1), right = FALSE))

## get sample size ----
n_figure_1 <- nrow(e_data_for_menarche)

##Figure 1
par(mfrow = c(1,1))



# Calculating mean and standard error for each age group
age_summary <- e_data_for_menarche %>%
  group_by(age_group) %>%
  summarise(
    mean_e_at_age_years = mean(e_conc_ug, na.rm = TRUE),
    se_e_at_age_years = sd(e_conc_ug, na.rm = TRUE) / sqrt(n())
  ) %>%
  ungroup()

print("Grouping ages into years")
ggplot(data = age_summary, aes(x = as.factor(age_group), y = mean_e_at_age_years)) + 
  geom_smooth() + 
  geom_point() +
  geom_vline(xintercept = mean_age_above_threshold, linetype = "dashed", color = "red") +
  geom_hline(yintercept = calculated_treshold, linetype = "dashed", color = "skyblue") +
  geom_errorbar(aes(ymin = mean_e_at_age_years - se_e_at_age_years, ymax = mean_e_at_age_years + se_e_at_age_years), width = 0.2) +
  geom_rect(aes(xmin = mean_age_above_threshold - 1.96 * se_age_above_threshold, xmax = mean_age_above_threshold + 1.96 * se_age_above_threshold, 
                ymin = -Inf, ymax = Inf), alpha = 0.05, fill = "pink") +
  ggtitle(paste0("Average estradiol by age at sample (N = ", n_figure_1, " samples)")) +
  xlab("Age at sample (years)") +
  ylab("Fecal estradiol (ug/g)")

print("Putting all data points and applying a smoothing function")
ggplot(data = e_data_for_menarche, aes(x = age_at_sample, y = e_conc_ug)) + 
  geom_smooth()+
  geom_point()+
  geom_vline(xintercept = mean_age_above_threshold, linetype = "dashed", color = "red") +
  geom_hline(yintercept = calculated_treshold, linetype = "dashed", color = "skyblue") +
  # geom_errorbar(aes(ymin = mean_e_at_age_years - se_e_at_age_years, ymax = mean_e_at_age_years + se_e_at_age_years), width = 0.2) +
  geom_rect(aes(xmin = mean_age_above_threshold - 1.96 * se_age_above_threshold, xmax = mean_age_above_threshold + 1.96 * se_age_above_threshold, 
                ymin = -Inf, ymax = Inf), alpha = 0.01, fill = "pink") +
  ggtitle(paste0("Average estradiol by age at sample (N = ", n_figure_1, " samples)")) +
  xlab("Age at sample (years)") +
  ylab("Fecal estradiol (ug/g)")

print("Showing all datapoints with box plots for age-year groupings and smoothing")
ggplot(data = e_data_for_menarche, aes(x = as.factor(floor(age_at_sample+1)), y = e_conc_ug)) +
  geom_point(data = e_data_for_menarche, aes(x = age_at_sample, y = e_conc_ug),
             alpha = 0.2, color = "green") +
    geom_boxplot() +
  geom_smooth(data = age_summary, aes(x = as.numeric(age_group), y = mean_e_at_age_years), method = "loess", color = "blue") +
  geom_vline(xintercept = mean_age_above_threshold, linetype = "dashed", color = "red") +
  geom_hline(yintercept = calculated_treshold, linetype = "dashed", color = "skyblue") +
  geom_rect(aes(xmin = mean_age_above_threshold - 1.96 * se_age_above_threshold, xmax = mean_age_above_threshold + 1.96 * se_age_above_threshold, 
                ymin = -Inf, ymax = Inf), alpha = 0.01, fill = "pink") +
  ggtitle(paste0("Estradiol levels by age at sample (N = ", n_figure_1, " samples)")) +
  xlab("Age at sample (years)") +
  ylab("Fecal estradiol (ug/g)") +
  scale_x_discrete(labels = c("0-1", "1-2", "2-3", "3-4", "4-5", "5-6", "6-7", "7-8", "8-9", "9-10"))


# ggplot(data = e_data |> 
#          filter(age_at_sample<10) |> 
#          select(age_at_sample_years, mean_e_at_age_years, se_e_at_age_years) |> 
#          distinct(), aes(x = age_at_sample_years, y = mean_e_at_age_years)) + 
#   geom_smooth() + 
#   geom_point()+
#   geom_vline(xintercept = mean_age_above_threshold, linetype = "dashed", color = "red") +
#   geom_hline(yintercept = calculated_treshold, linetype = "dashed", color = "skyblue") +
#   geom_errorbar(aes(ymin = mean_e_at_age_years - se_e_at_age_years, ymax = mean_e_at_age_years + se_e_at_age_years), width = 0.2) +
#   geom_rect(aes(xmin = mean_age_above_threshold - 1.96*se_age_at_menarche, xmax = mean_age_above_threshold + 1.96*se_age_at_menarche, 
#                 ymin = -Inf, ymax = Inf), alpha = 0.05, fill = "pink") +
#   ggtitle(paste0("Average estradiol by age at sample (N = ", n_figure_1, " samples)"))+
#   xlab("Age at sample")+
#   ylab("Fecal estradiol (ug/g)")
```

\
\

Interpreting the graph:

-   The black circles and error bars represent the average age fE2 reading (with standard error) for each age-year for the population

-   The blue line represents the results of a Loess regression, a non-parametric approach to modeling the relationship between x and y. The gray shading around the blue line is the error of the estimate.

-   The dashed light blue line is the calculated prepubertal fE threshold (y = 2.845277 ug/g)

-   The dashed red line is the calculated average age at puberty (x = 4.641429 years)

-   The shaded area around the dashed red line is the 95% CI for the average age at puberty estimate (3.22 ± 1.96\* SE)

## Age at menarche on the individual-level

### Plot of fE2 as a function of age: individual-level

NOTE: I haven't yet calculated individual E thresholds/ages at menarche because I think we are going to want a different approach than mean + 2sd....still working on that.

I am pulling data on females that we have good E coverage for before and after potential menarche...I'm looking for IDs with samples starting at least by age 3. At least 25 samples per ID. That leaves us with the following:

```{r,class.source = 'fold-hide', warning = FALSE}
par(mfrow = c(1,1))

# Panel of all individual females who have good E coverage before and after menarche
females_for_menarche_panel<-e_data |> 
  group_by(ind_id) |> 
  filter(min(age_at_sample)<3) |>
  mutate(count = n()) |> 
  ungroup() |> 
  select(ind_id,count) |> 
  arrange(desc(count)) |> 
  distinct() |> 
  filter(count>25) |> 
  pull(ind_id)

females_for_menarche_panel_df<- e_data |> 
  filter(ind_id %in% females_for_menarche_panel) |>
  filter(age_at_sample < 8) |>
  select(ind_id, age_at_sample, sample_id, date, month, time, hour, e_conc_ug, is_pregnant) |> 
  arrange(ind_id, age_at_sample) |> 
  mutate(date = as.Date(date))

n_figure_2 <-nrow(females_for_menarche_panel_df)

# facet wrap plots with geom_smooth
ggplot(data = females_for_menarche_panel_df, aes(x = age_at_sample, y = e_conc_ug, color = ind_id)) +
  geom_point()+
  facet_wrap(~ ind_id)+
  geom_smooth()+
  ggtitle(paste0("Estradiol by age at sample, separated by individual (n = ", n_figure_2, " samples)"))+
  xlab("Age at sample")+
  ylab("Fecal estradiol (ug/g)")+
  theme(legend.position = "none")

# Loop through each individual and create a ggplot for each
for (individual in females_for_menarche_panel) {
  individual_data <- females_for_menarche_panel_df |> 
    filter(ind_id == individual)
  
  # Get the age at which estradiol first exceeds the threshold for this individual
  first_above_threshold_age_by_id <- first_above_threshold |> 
    filter(ind_id == individual) |> 
    pull(first_age_above_threshold)
  
  print(paste0("Age at menarche for ", individual,": ", first_above_threshold_age_by_id))
  
  # Get duckface dates
  duckface_ages <- duckface_data %>%
    filter(ind_id == individual) %>%
    pull(age_at_sample)
  
  
  # Create the plot
  p <- ggplot(data = individual_data, aes(x = age_at_sample, y = e_conc_ug)) +
    geom_point() +
    geom_line(color = "blue") +
    geom_vline(xintercept = first_above_threshold_age_by_id, 
               linetype = "dashed", color = "red") +
     geom_vline(xintercept = as.numeric(duckface_ages), linetype = "solid", color = "purple") +  # Add vertical lines for duckface dates
    ggtitle(paste0("Estradiol by age at sample for individual ", individual, " (n = ", nrow(individual_data), " samples)")) +
    xlab("Age at sample") +
    ylab("Fecal estradiol (ug/g)") +
    theme(legend.position = "none")
  
  print(p)
}
```

### Modeling E and age

#### Step 1. Assessing normality

```{r}
shapiro.test(females_for_menarche_panel_df$e_conc_ug) #nope
 
shapiro.test(log(females_for_menarche_panel_df$e_conc_ug)) # log transformation does the trick!
```

Log transformation does the trick! Data is now normal for our young female panel.

#### Step 2. Modeling E2 and age

I am including individual ID as a random effect. In unadjusted model, fixed effects are month and time of day (accounting for seasonality and diurnal variation) with individual ID included as a random effect. Any other things to control for?\
\
The adjusted model includes age as a fixed effect. Age improves the model fit!

```{r}
unadjusted_e_model <- lmer(log(e_conc_ug) ~  + month + hour+ (1 | ind_id), data = females_for_menarche_panel_df)

adjusted_e_model <- lmer(log(e_conc_ug) ~ age_at_sample + month + hour+ (1 | ind_id), data = females_for_menarche_panel_df)

anova(unadjusted_e_model, adjusted_e_model) # adjusted_e_model fits better

summary(adjusted_e_model)

#checking for multicollinearity -- low (~ 1)
vif(adjusted_e_model)
print("No multicollinearity -- yay!")
```

#### Step 3. Checking model diagnostics

Tldr: model passes diagnostics—yay

```{r}
# checking model diagnostics
plot(adjusted_e_model) #resids vs leverage -- pretty good
lattice::qqmath(adjusted_e_model) ## qq norm plot -- good
plot(adjusted_e_model, # scale location plot -- pretty good
     sqrt(abs(resid(.)))~fitted(.),
     type=c("p","smooth"), col.line=1)
```

## Menarche and Duck Face

NOTE: I am holding off on reporting on this until we have the updated behavioral data containing group scan observations of duck face....planning to adapt methods from [Sen et al. 2022](https://www.sciencedirect.com/science/article/pii/S0018506X22001581?casa_token=6acu1ntMv9UAAAAA:DE2kCyrQg3ooVPhuqtaTpwmtINgwtDO-ruFqdxLt7dMWHL2i4Xey0m_wzLH_ukl9rW56jXeZ8g#t0010)

# **Cycling E**

## Establish Cycle Length

```{r warning = FALSE,  message=FALSE}
library(dplyr)
library(ggplot2)
library(zoo) # For rollapply
# Panel of all individual females who have good E coverage before and after menarche
females_for_cycling_panel<-e_data |> 
  group_by(ind_id) |> 
  filter(min(age_at_sample)>3) |> 
  mutate(count = n()) |> 
  ungroup() |> 
  select(ind_id,count) |> 
  arrange(desc(count)) |> 
  distinct() |> 
  filter(count>25) |> 
  pull(ind_id)

females_for_cycling_panel_df<- e_data |> 
  filter(ind_id %in% females_for_cycling_panel) |>
  mutate(date = as.Date(date)) |> 
  filter(date > as.Date("2021-12-31")) |> 
  select(ind_id, age_at_sample, sample_id, date, month, time, hour, e_conc_ug, is_pregnant) |> 
  arrange(ind_id, age_at_sample) 
n_figure_2 <-nrow(females_for_cycling_panel_df)

create_continuous_segments <- function(df) {
  df %>%
    arrange(date) %>%
    mutate(segment = cumsum(is_pregnant == "P") + 1) %>%
    group_by(segment) %>%
    filter(any(is_pregnant == "NP")) %>%
    ungroup() %>%
    mutate(segment = cumsum(is_pregnant == "P") + 1) %>%
    filter(is_pregnant == "NP")
}

# Function to find local peaks
find_peaks <- function(df) {
  df %>%
    arrange(date) %>%
    mutate(peak = rollapply(e_conc_ug, width = 3, FUN = function(x) x[2] == max(x), fill = NA, align = 'center')) %>%
    filter(peak == TRUE) %>%
    select(ind_id, date, e_conc_ug, is_pregnant)
}

# Function to calculate cycle lengths
calculate_cycle_lengths <- function(df) {
  df <- df |> 
    filter(is_pregnant == "NP")
  
  peaks <- find_peaks(df)
  peaks <- peaks %>%
    arrange(date) %>%
    mutate(next_date = lead(date)) %>%
    mutate(cycle_length = as.numeric(difftime(next_date, date, units = "days"))) %>%
    filter(!is.na(cycle_length))
  
  # Filter to only include cycle lengths within 45 days
  valid_cycles <- peaks %>%
    filter(cycle_length <= 45)
  
  # Calculate average cycle length
  avg_cycle_length <- valid_cycles %>%
    summarise(avg_cycle_length = mean(cycle_length, na.rm = TRUE))
  
  return(avg_cycle_length)
}

# Loop through each individual and create a ggplot for each
for (individual in females_for_cycling_panel) {
  individual_data <- females_for_cycling_panel_df %>%
    filter(ind_id == individual)
  
  # Calculate cycle lengths for this individual
  cycle_lengths <- calculate_cycle_lengths(individual_data)
  
  # Create continuous segments based on NP periods
  segmented_data <- create_continuous_segments(individual_data)
  
 p<-ggplot(data = individual_data, aes(x = date, y = e_conc_ug)) +
    geom_point(aes(color = is_pregnant)) +
    ggtitle(paste0("Estradiol by date for individual ", individual, " (n = ", nrow(individual_data), " samples)")) +
    xlab("Date") +
    ylab("Fecal estradiol (ug/g)") +
    theme(legend.position = "none") +
    geom_line(data = segmented_data, aes(group = segment), color = "blue") +  # Plot lines for continuous NP segments
       geom_vline(xintercept = as.Date(cycle_lengths$avg_cycle_length), linetype = "dashed", color = "green")

  print(p)
  
  # Print cycle lengths for this individual
  print(paste0("Average cycle length for individual ", individual, ": ", cycle_lengths$avg_cycle_length, " days"))
}

```

## Visualizing cycle E and Duckface

Purple lines = observed duck face

```{r warning = FALSE, message=FALSE}

for (individual in females_for_cycling_panel) {
  individual_data <- females_for_cycling_panel_df %>%
    filter(ind_id == individual)
  
  # Calculate cycle lengths for this individual
  cycle_lengths <- calculate_cycle_lengths(individual_data)
  
  # Create continuous segments based on NP periods
  segmented_data <- create_continuous_segments(individual_data)
  
  # Get duckface dates for this individual
  duckface_dates <- duckface_data %>%
    filter(ind_id == individual) %>%
    pull(date)
  
  
  p <- ggplot(data = individual_data, aes(x = date, y = e_conc_ug)) +
    geom_point(aes(color = is_pregnant)) +
    ggtitle(paste0("Estradiol by date for individual ", individual, " (n = ", nrow(individual_data), " samples)")) +
    xlab("Date") +
    ylab("Fecal estradiol (ug/g)") +
    theme(legend.position = "none") +
    geom_line(data = segmented_data, aes(group = segment), color = "blue") +  # Plot lines for continuous NP segments
    geom_vline(xintercept = as.numeric(duckface_dates), linetype = "solid", color = "purple") +  # Add vertical lines for duckface dates
    geom_vline(xintercept = as.Date(cycle_lengths$avg_cycle_length), linetype = "dashed", color = "green")

  print(p)
  
  # Print cycle lengths for this individual
  print(paste0("Average cycle length for individual ", individual, ": ", cycle_lengths$avg_cycle_length, " days"))
}
```

### Testing association between duckface and E Peaks

```{r}

# Combine data from all individuals
all_peaks <- data.frame()
all_duckface_events <- data.frame()

for (individual in females_for_cycling_panel) {
  individual_data <- females_for_cycling_panel_df %>%
    filter(ind_id == individual)
  
  # Find peaks for this individual
  peaks <- find_peaks(individual_data)
  all_peaks <- rbind(all_peaks, peaks)
  
  # Get duckface dates for this individual
  duckface_dates <- duckface_data %>%
    filter(ind_id == individual) %>%
    pull(date)
  
  duckface_events <- data.frame(ind_id = individual, date = duckface_dates)
  all_duckface_events <- rbind(all_duckface_events, duckface_events)
}

# Check for duckface events within ±4 days of peak estradiol dates
all_peaks <- all_peaks %>%
  mutate(duckface_within_4_days = sapply(date, function(peak_date) {
    any(abs(difftime(all_duckface_events$date[all_duckface_events$ind_id == ind_id], peak_date, units = "days")) <= 4)
  }))

# Summarize results
summary_table <- all_peaks %>%
  group_by(ind_id) %>%
  summarise(
    total_peaks = n(),
    peaks_with_duckface = sum(duckface_within_4_days),
    proportion_with_duckface = mean(duckface_within_4_days)
  )

# Perform statistical test
# Create a contingency table
contingency_table <- table(all_peaks$duckface_within_4_days)
# Perform chi-squared test
chi_squared_test <- chisq.test(contingency_table)

# Print results
print(summary_table)
print(chi_squared_test)

```

# **Pregnancy E**

## Age at first birth

Could you please share the estimated/confirmed ages of all of our IDs? I can't really assign this easily (unless I work backwards from age at sample info....). I know I had made a Google Sheet at some point with this information, but I no longer have access. Thanks!

## Assigning gestation lengths

```{r}
e_data <- e_data %>%
  mutate(across(c(B2,B3,B4), ~ na_if(.x, " ")))

e_data$B1 <- as.Date(e_data$B1)
e_data$B2 <- as.Date(e_data$B2)
e_data$B3 <- as.Date(e_data$B3)
e_data$B4 <- as.Date(e_data$B4)

e_data$date <- as.Date(e_data$date)
# # Calculate the difference in months
# 
gestation_df_long <- e_data %>%
   pivot_longer(cols = c(B1, B2, B3, B4), names_to = "birth_number", values_to = "birthing_date")

gestation_df <- gestation_df_long |> 
  mutate(months_from_birth = interval(date, birthing_date) / months(1)) |> 
  select(ind_id, age_at_sample, month, months_from_birth, e_conc_ug, birth_number)

gestating_df<- gestation_df |> 
  filter(months_from_birth >= 0 & months_from_birth <= 9) |> 
  mutate(months_from_birth = -1*months_from_birth)

gestating_df_females<-gestating_df |> 
  distinct(ind_id) |> 
  pull(ind_id) 

not_gestating_df<-gestation_df |> 
  filter(is.na(months_from_birth) | months_from_birth > 9)  |> 
  filter(age_at_sample > 6) |> 
  group_by(ind_id) |> 
  summarise(non_gestating_e2 = mean(e_conc_ug))

  
# Plot the data for each individual
for (individual in gestating_df_females) {
  individual_data <- gestating_df %>%
    filter(ind_id == individual)
  
  # Establish baseline E2 for the individual
  baseline_e <- not_gestating_df |> 
    filter(ind_id == individual) |>
    pull(non_gestating_e2)
  
  p <- ggplot(data = individual_data, aes(x = months_from_birth, y = e_conc_ug, color = birth_number, linetype = birth_number)) +
    geom_point() +
    geom_line() +
    geom_hline(yintercept = baseline_e, linetype = "dashed", color = "green") +
    ggtitle(paste0("Estradiol and gestation for ", individual, " (n = ", nrow(individual_data), " samples)")) +
    xlab("Months from birth") +
    ylab("Fecal estradiol (ug/g)") +
    ylim(0, 150) +
    xlim(-9, 0) +
    theme(legend.position = "right") +
    labs(color = "Birth Number", linetype = "Birth Number")

  print(p)
}
```

## E across trimesters

**QUESTION**: I see some individuals are marekd as pregnant (e..g, `is_pregnant` = "P"), but the `preg_trim` column is blank. What do I do about this?

For now, I'm excluding pregnant individuals with missing trimester info.

### Step 1. Preparing dataset for pregnant individuals with known trimester info

```{r}
pregnant_df<-e_data |> 
  select(ind_id, month, hour, age_at_sample, e_conc_ug, preg_trim, pregnancy_num, is_pregnant) |> 
  filter(is_pregnant == "P") |> 
  filter(preg_trim !="")

n_pregnant_df <- nrow(pregnant_df)

```

### Step 2. Data visualization

```{r}
par(mfrow = c(1,1))
ggplot(data = pregnant_df, aes(x = preg_trim, y = e_conc_ug))+
  geom_boxplot()+
  geom_jitter(alpha = 0.1)+
  ggtitle(paste("Estradiol and gestation (N = ", n_pregnant_df, "samples)")) +
  xlab("Pregnancy trimester")+
  ylab("Fecal estradiol (ug/g)")
```

### Step 3. Assessing normality

Data is normal following log transformation

```{r}
shapiro.test(pregnant_df$e_conc_ug)
shapiro.test(log(pregnant_df$e_conc_ug))
```

### Step 4. ANOVA and Tukey's HSD

Among trimesters, we only see that T2 is significantly lower than T3; no significant difference between T1-T2 or T1-T3.

```{r}
anova_result_pregancy_e <- aov(log(e_conc_ug) ~ preg_trim, data = pregnant_df)

summary(anova_result_pregancy_e)
tukey_results_pregnancy_e <- TukeyHSD(anova_result_pregancy_e)
print(tukey_results_pregnancy_e)
```

### Step 5: Modeling E2 across pregnancy

```{r}
## what other control variables should I include?
unadjusted_pregnancy_model <- lmer(log(e_conc_ug) ~ age_at_sample + month+ hour+ (1 | ind_id), data = pregnant_df)

adjusted_pregnancy_model <- lmer(log(e_conc_ug) ~ age_at_sample + month + hour+preg_trim + (1 | ind_id), data = pregnant_df)

anova(unadjusted_pregnancy_model, adjusted_pregnancy_model) # adjusted model is better fit--pregnancy predicts E

summary(adjusted_pregnancy_model)
```

### Step 6: Checking model diagnostics

Tldr: model passes diagnostics—yay

```{r}
#checking for multicollinearity -- low (~ 1)
vif(adjusted_pregnancy_model)

# checking model diagnostics
plot(adjusted_pregnancy_model) #resids vs leverage -- good
lattice::qqmath(adjusted_pregnancy_model) ## qq norm plot -- good...small tail on right
plot(adjusted_pregnancy_model, # scale location plot -- good
     sqrt(abs(resid(.)))~fitted(.),
     type=c("p","smooth"), col.line=1)
```

# **Lactation E** 

## Step 1. Preparing dataset for lactating vs not lactating

Only looking at adults (ages \>6)

```{r}
lactation_df <- e_data |> 
  select(ind_id, month, hour, age_at_sample, e_conc_ug, lactating) |> 
  filter(age_at_sample > 6)

n_lactation_df<-nrow(lactation_df)
```

### Step 2. Data visualization

```{r}
ggplot(data = lactation_df, aes(x = lactating, y = e_conc_ug))+
  geom_boxplot()+
  geom_jitter(alpha = 0.1)+
  ggtitle(paste("Estradiol and lactation (N = ", n_pregnant_df, "samples)")) +
  xlab("Lactating")+
  ylab("Fecal estradiol (ug/g)")
```

### Step 3. Assessing normality

Data is still not normal following log transformation

```{r}
shapiro.test(lactation_df$e_conc_ug)

shapiro.test(log(lactation_df$e_conc_ug))
```

### Step 4. Mann Whitney U Test

Data is not normal, so can't do ANOVA. Let's use a nonparametric approach to see if E is significantly lower during lactation....it is (W = 98014, p \< 0.001)

```{r}
wilcox.test(log(e_conc_ug) ~ lactating, data = lactation_df)
```

### Step 5. Modeling E2 and lactation

Technically, data should be normal for this. BUT I think our sample might be large enough that we're okay violating this assumption?

```{r}
## what other control variables should I include?
unadjusted_lactation_model <- lmer(log(e_conc_ug) ~ age_at_sample + month+ hour + (1 | ind_id), data = e_data)

adjusted_lactation_model <- lmer(log(e_conc_ug) ~ age_at_sample + month+ hour + lactating + (1 | ind_id), data = e_data)

anova(unadjusted_lactation_model,adjusted_lactation_model) # adjusted model is better fit--lactation predicts E

summary(adjusted_lactation_model) # lactation = lower E

```

### Step 6. Checking model diagnostics

Tldr: model passes diagnostics—yay

```{r}
#checking for multicollinearity -- low (~ 1)
vif(adjusted_lactation_model)

# checking model diagnostics
plot(adjusted_lactation_model) #resids vs leverage -- good
lattice::qqmath(adjusted_lactation_model) ## qq norm plot -- good...small tail on right
plot(adjusted_lactation_model, # scale location plot -- good
     sqrt(abs(resid(.)))~fitted(.),
     type=c("p","smooth"), col.line=1)
```

# **Male takeovers and E**

Can I please have access to these data?
